---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# `accrualPlot`


<!-- [![](https://www.r-pkg.org/badges/version/accrualPlot?color=green)](https://cran.r-project.org/package=accrualPlot)  -->
`r badger::badge_custom("dev version", as.character(packageVersion("accrualPlot")), "blue", "https://github.com/CTU-Bern/accrualPlot")`
[![Actions Status](https://github.com/CTU-Bern/presize/workflows/R-CMD-check/badge.svg)](https://github.com/CTU-Bern/presize/actions)
<!-- ![travis](https://travis-ci.com/CTU-Bern/presize.svg?branch=master) -->
<!-- [![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/CTU-Bern/presize?branch=master&svg=true)](https://ci.appveyor.com/project/CTU-Bern/presize) -->
<!-- [![codecov](https://codecov.io/github/CTU-Bern/accrualPlot/branch/master/graphs/badge.svg)](https://codecov.io/github/CTU-Bern/accrualPlot) -->

Accrual plots are an important tool when monitoring clinical trials. Some trials are terminated early due to low accrual, which is a waste of resources (including time). Assessing accrual rates can also be useful for planning analyses and estimating how long a trial needs to continue recruiting participants.  `accrualPlot` provides tools for such plots

## Installation

<!-- `accrualPlot` can be installed from CRAN in the usual manner: -->

```{r cran-installation, eval = FALSE, include=FALSE}
install.packages("presize")
```


You can install the development version of `accrualPlot` from github with:

```{r gh-installation, eval = FALSE}
# install.packages("remotes")
remotes::install_github("CTU-Bern/accrualPlot")
```

Note that `remotes` treats any warnings (e.g. that a certain package was built under a different version of R) as errors. If you see such an error, run the following line and try again:

```{r remotes-error, eval = FALSE}
Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS = "true")
```

## Overview
The first step to using `accrualPlot` is to create an accrual dataframe. This is simply a dataframe with a counts of participants included per day.


```{r}
# load package
library(accrualPlot)

# generate some data
set.seed(1234)
x <- as.Date("2020-12-07") + sample(c(-20:20), 50, replace = TRUE)

df <- accrual_create_df(x)

```

Cumulative recruitment
```{r}
plot(df)
# accrual_plot_cum(df)
```

Recruitment per day
```{r}
plot(df, which = "abs", unit = "day")
# accrual_plot_abs(df, unit = "day")
```

If we know that we started recruiting on the 1st November, we can add this information.
```{r}
# accrual_plot_cum(df, start_date = as.Date("2020-11-01"))
plot(df, start_date = as.Date("2020-11-01"))
```

Multiple sites can also be depicted...
```{r}
site <- sample(1:3, 50, replace = TRUE)
df2 <- accrual_create_df(x, by = site)
# accrual_plot_cum(df2)
plot(df2)
```

It is also possible to predict the time point at which a certain number of participants has been recruited (for estimating when a study will be complete). If we want to recruit a total of 75 participants, we can put that in the `target` option. 

```{r}
accrual_plot_predict(df, target = 75)
# accrual_plot_predict(df2, target = 75) # does not seem to work
plot(df, "predict", target = 75)
```

Table of recruitment, with or without a descriptive header.
```{r}
# accrual_table(df) 
summary(df) 
summary(df2) 
summary(df2, header = FALSE) 
```



